---
title: "HW 6"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter your name and EID here:

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

We will use the packages `tidyverse`, `factoextra`, and `cluster` for this assignment.

```{r, message = FALSE}
# Load packages
library(tidyverse)
library(factoextra)
library(cluster)
```

------------------------------------------------------------------------

## Question 1: (2 pts)

The dataset for this homework comes from the article:

*Tsuzuku N, Kohno N. 2020. The oldest record of the Steller sea lion Eumetopias jubatus (Schreber, 1776) from the early Pleistocene of the North Pacific. <https://doi.org/10.7717/peerj.9709>*

Read the ***Abstract*** of the article and the section called *Results of Morphometric Analyses*. What was the goal of this study and what was the main finding?

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 2: (1 pt)

Under the supplemental information, I retrieved the data from a word document into a `.csv` document. Read in the dataset. 

```{r}
# download data from GitHub
sealions <- read_csv("Sealions.csv")
```

How many rows and how many columns are in this dataset? What does a row represent? What does a column represent?

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 3: (1 pt)

Before we can analyze the data, let's do some cleaning. Using a combination of the `select()`, `where()`, and a predicate function like `is.character()` we can scan through all the columns of the dataset and see which columns are of character type.

```{r}
## Select all of the columns that have character type
sealions |> 
    select(where(is.character))
```

When importing this dataset into RStudio, which variables were considered numeric? Use the `is.numeric()` predicate function here. 

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

Take a look at one of the character columns. Why are some measurements not considered as numeric? 

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 4: (1 pt)

The functions `mutate()` and `across()`, when used together, can make changes across a range of columns in a data frame. 

Using `mutate()` and `across()`, replace all `-` in the dataset with the missing values *NA* and then then make sure all measurements are defined as numeric variables with `as.numeric()`. The first part of the code replaces the "-" with NA values. Write the second part to coerce all of the columns (except for the ID column!) to be numeric. Make sure to overwrite the dataset `sealions`. 

NOTE: Look at the examples in `?across` to get a sense of how to use the `across()` function.

```{r}
# your code goes below (make sure to edit comment)
sealions |> 
    ## Replace all "-" with NA
    mutate(across(where(is.character), ~ na_if(.x, "-"))) |> 
    ## Coerce all columns (except for ID) to be numeric
```

What is the mean rostral tip of mandible `C`?

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 5: (2 pts)

You are given the code in this question. But what does the code do? Write comments.

```{r}
sealions <- sealions |> 
  ## 
  select_if(!is.na(sealions[51,])) |> 
  ## 
  na.omit()
```

How many columns and how many rows are remaining in this dataset?

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 6: (2 pts)

Use the `separate()` function on `sealions` to split the `ID` variable into two variables `species` and `sex`. The article states that the fossil specimen has to be male. Replace the missing value of `sex` for the fossil specimen GKZ-N 00001. 

*Hint: in the `ID` variable, what symbol separates the species from sex?* 

*Hint: You could use the functions `mutate()` and `replace_na()`.* 

*Hint: When separating by a special character like `[`, you need to escape it using a `\\`*

Save the resulting dataset as `sealions_clean`.

```{r}
# your code goes below (make sure to edit comment)
 
```

How many sealions are male/female?

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 7: (1 pt)

Using `tidyverse` functions, only keep numeric variables and scale each numeric variable. Save the resulting dataset as `sealions_num`. What should the mean of the scaled variable of the rostral tip of mandible C be?

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 8: (2 pts)

Let's  perform PCA on the measurements available for the fossil specimen GKZ-N 00001. Using the function `prcomp()`, calculate the principal components (PCs) for the scaled data, `sealions_num`, obtained in the previous question. Construct a scree plot with the function `fviz_eig()` from the package `factoextra`. What is the cumulative percentage of explained variance for PC1 and PC2? 

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 9: (2 pts)

How many *known species* are there in `sealions_clean`? Therefore, how many clusters should we look for to identify what species GKZ-N 00001 most likely belongs to? 

```{r}
# your code goes below (make sure to edit comment)
 
```

**Your answer goes here. Write sentences in bold.**

Try using the PAM clustering algorithm using the `pam()` function in R (it is similar to K-means). Perform the PAM clustering algorithm on `sealions_num`, run the PAM clustering algorithm with 3 clusters and save the results to an object named `pam_results`.

```{r}
# your code goes below (make sure to edit comment)
 
```

------------------------------------------------------------------------

## Question 10: (2 pts)

Represent the clusters along the first two principal components and specify to shape the observations by their `species` in the aesthetics. *Note: you can either use `ggplot` or `fviz_cluster()`.* 

```{r}
# your code goes below (make sure to edit comment)
fviz_cluster(pam_results, geom = NULL) +  ## Add geoms to complete the plot
```

The fossil specimen GKZ-N 00001 appears to be close to which species? 

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Question 11: (2 pts)

Putting it all together. Reflect on and summarize in 1-2 sentences the different steps taken through this assignment. Compare your conclusions to the findings discussed by the researchers in the article (cite their findings).

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

## Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```
